---
title: "Reproduced analysis of FARS data"
output: pdf_document 
--- 

```{r, warning=FALSE,echo=TRUE, message=FALSE}
library(dplyr)
library(tidyverse)
library(readr) 
library(tufte) 
library(tibble) 
library(broom)
library(tidyr)
library(purrr)
library(foreign)
library(ggplot2)
library(ggthemes)
library(stats)
library(knitr)

load("../data/fars_data.RData")
source("../R/fars_functions.R")
```


## **Results**

#### **Percentages of drivers testing positive by drug type, sex, and year group**

```{r, warning=FALSE,echo=TRUE, message=FALSE}

clean_fars %>%
  mutate(year_cat = cut(year, breaks = c(1999, 2002, 2006, 2010),
                        labels = c("1999-2002", "2003-2006",
                                   "2007-2010"),
                        include.lowest = TRUE, right = TRUE)) %>%
  filter(!is.na(sex)) %>%
  group_by(drug_type, sex, year_cat) %>%
  summarize(n_non_missing = sum(!is.na(positive_for_drug)),
            positive_test = sum(positive_for_drug, na.rm = TRUE),
            perc_positive = round(100 * positive_test / n_non_missing, 1)) %>%
  select(drug_type, sex, year_cat, perc_positive) %>%
  unite(sex_year_cat, sex, year_cat) %>%
  spread(sex_year_cat, perc_positive) %>%
  knitr::kable(col.names = c("Drug type", "F 1999-2002",
                             "F 2003-2006", "F 2007-2010",
                             "M 1999-2002", "M 2003-2006",
                             "M 2007-2010"))

```

#### **Figure 1: Prevalence of nonalcohol drugs in fatally injured drivers by year and age group**

```{r,warning=FALSE,echo=TRUE, message=FALSE, fig.width=7, fig.height=3.75}
clean_one <- clean_fars %>% 
  filter(drug_type != "Alcohol") %>%
  filter(!is.na(agecat)) %>%
  group_by(unique_id, year, agecat) %>% 
  summarize(positive_tests = sum(positive_for_drug, na.rm = TRUE), positive = any(positive_tests > 0)) %>% 
  ungroup() %>% 
  group_by(year, agecat) %>% 
  summarize(total_tests = length(positive), positive = sum(positive), percent_positive = round(100 * positive/total_tests, 1)) %>%
  ggplot(aes(x = year, y = percent_positive, position = agecat)) +
  geom_point(aes(shape = agecat)) +
  geom_line() +
  labs(x = "Year", y = "Positive for Nonalcohol Drugs, %", shape = "Age") +
  theme_few() +
  scale_y_continuous(limits = c(0, 35))
  
clean_one
```
   

\newpage


   
#### **Figure 2: Prevalence of nonalcohol drugs in fatally injured drivers by year and drug type**
```{r,warning=FALSE,echo=TRUE, message=FALSE,fig.width=7, fig.height=3.75}
clean_two <- clean_fars %>% 
  filter(drug_type != "Alcohol") %>% 
  filter(!is.na(positive_for_drug)) %>% 
  group_by(drug_type, year) %>% 
  summarize(n_not_NA = sum(!is.na(positive_for_drug)),
            positive_tests = sum(positive_for_drug, na.rm = TRUE),
            percent_positive = round(100 * positive_tests/n_not_NA,1)) %>% 
  ggplot(aes(x = year, y = percent_positive, position = drug_type)) +
  geom_point(aes(shape = drug_type)) +
  geom_line() +
  labs(x = "Year", y = "Positive for Drugs, %", shape = "Drug type") +
  theme_few() +
  scale_y_continuous(breaks = c(0, 4, 8, 12))
clean_two
```

#### **Figure 3: Prevalence of cannabinoid drugs in fatally injured drivers by year and age group**

```{r,warning=FALSE,echo=TRUE, message=FALSE,fig.width=7, fig.height=3.75}

clean_three <- clean_fars %>%
  filter(drug_type == "Cannabinoid") %>% 
  filter(!is.na(agecat)) %>% 
  group_by(agecat, year) %>% 
  summarize(n_not_NA = sum(!is.na(positive_for_drug)),
            positive_tests = sum(positive_for_drug, na.rm = TRUE),
            percent_positive = round(100 * positive_tests/n_not_NA,1)) %>%
  ggplot(aes(x = year, y = percent_positive, position = agecat)) +
  geom_line() +
  geom_point(aes(shape = agecat)) +
  labs(x = "Year", y = "Positive for Cannabinoid, %", shape = "Age") +
  theme_few()
clean_three
```

\newpage

#### **Prevalence of drugs in fatally injured drivers for 1999 and 2010 by drug type**


```{r, warning=FALSE,echo=TRUE, message=FALSE}
perc_cis <- function(x = x, n = n){
  library(scales) 
  
  proportion <-  x/n
  est_of_stand <-  sqrt((proportion*(1 - proportion))/n)
  low_CI <- proportion-(1.96*est_of_stand)
  upp_CI <- proportion+(1.96*est_of_stand)

prop <- percent(proportion)
low_conf <- percent(low_CI)
upp_conf <- percent(upp_CI)


perc_cis_vec <- paste(prop, " ", "(", low_conf, ",", " ",upp_conf, ")", sep = "")
return(perc_cis_vec)
}

prev <- clean_fars %>% 
  filter(year %in% c("1999","2010")) %>% 
  group_by(year, drug_type) %>% 
  summarize(positive = sum(positive_for_drug, na.rm = TRUE), trials = sum(!is.na(positive_for_drug)))

prev_2 <- data.frame(prevalence = perc_cis(x = prev$positive, n = prev$trials))

prev <- rowid_to_column(as.data.frame(prev, var = "rowid"))
prev_2 <- rowid_to_column(as.data.frame(prev_2, var = "rowid"))

prevalence_table <- prev %>% 
  full_join(prev_2, by = "rowid") %>% 
  select(prevalence, drug_type, year) %>% 
  spread(key = year, value = prevalence) %>% 
  rename("Drug type" = drug_type)

kable(prevalence_table) 


```



#### **Statistics for testing for trend in prevalence of drugs over study years by drug type using Cochran-Armitage trend test**

```{r, warning=FALSE,echo=TRUE, message=FALSE}

test_trend_ca <- function(drug, data = clean_fars) {
  if(drug == "Nonalcohol"){
    clean <- clean_fars %>%
      filter(drug_type != "Alcohol") %>%
      group_by(unique_id, year) %>% 
      summarize(positive_test = sum(positive_for_drug, na.rm = TRUE), 
                positive = any(positive_test > 0),
                total_tests = length(!is.na(positive_for_drug))) %>% 
      ungroup() %>% 
      group_by(year) %>%
      summarize(total_tests = sum(total_tests), positive = sum(positive)) 
  } else{
    clean <- clean_fars %>%
      filter(drug_type == drug) %>%
      group_by(year) %>%
      summarize(positive = sum(positive_for_drug, na.rm = TRUE),
                total_tests = sum(!is.na(positive_for_drug)))
  }
  ca_drug_type <- prop.trend.test(x = clean$positive,
                                  n = clean$total_tests)
  Z <-  sqrt(ca_drug_type$statistic)
  p.value <- ca_drug_type$p.value
  
  Z <- round(Z, digits = 1)
    p.value <- round(p.value, digits = 3)
  
  test_trend_table <- tibble::tibble(Z, p.value)
  
  test_trend <- test_trend_table %>% 
    mutate(Z = as.numeric(Z)) %>% 
    mutate(p.value = as.numeric(p.value))
  
  return(test_trend)
}

drug_list <- c("Alcohol", "Nonalcohol", "Narcotic", "Depressant",
"Stimulant", "Cannabinoid", "Other")
drug_trend_tests_ca <- lapply(drug_list, test_trend_ca)
drug_trend_tests_ca <- dplyr::bind_rows(drug_trend_tests_ca) %>%
dplyr::mutate(drug = drug_list) %>%
dplyr::select(drug, Z, p.value)
drug_trend_tests_ca %>% knitr::kable()
  

```

#### **Statistics for testing for trend in prevalence of drugs over study years by drug type using Wald test of logistic regression coefficient for “year”**

```{r, warning=FALSE,echo=TRUE, message=FALSE}

test_trend_log_reg <- function(drug, data = clean_fars) {
    if(drug == "Nonalcohol"){
      to_test <- clean_fars %>%
        filter(!is.na(drug_type)) %>% 
        filter(drug_type != "Alcohol") %>% 
        group_by(unique_id, year) %>% 
        summarize(positive_for_drug = any(positive_for_drug))
      
} else{
      to_test <- clean_fars %>%
        filter(!is.na(drug_type)) %>%
        filter(drug_type == drug)
}
    log_reg <- glm(positive_for_drug ~ year, data = to_test,
                   family = binomial(link = "logit"))
    log_reg_clean <- slice(tidy(log_reg), 2)
    Z <- log_reg_clean$statistic
    p.value <- log_reg_clean$p.value
    
    Z <- round(Z, digits = 1)
    p.value <- round(p.value, digits = 3)
    
    log_reg_table <- tibble::tibble(Z, p.value)
    
    log_reg_trend <- log_reg_table %>% 
      mutate(Z = as.numeric(Z)) %>% 
      mutate(p.value = as.numeric(p.value))
    
   return(log_reg_trend)
}


drug_list <- c("Alcohol", "Nonalcohol", "Narcotic", "Depressant",
"Stimulant", "Cannabinoid", "Other")
drug_trend_tests_log_reg <- lapply(drug_list, test_trend_log_reg)
drug_trend_tests_log_reg <- dplyr::bind_rows(drug_trend_tests_log_reg) %>%
dplyr::mutate(drug = drug_list) %>%
dplyr::select(drug, Z, p.value)
drug_trend_tests_log_reg %>% knitr::kable()

```





